### 2012-12-19 doc/description.tex
a lot of fixes,  addProperty with data

$$
Application = <Class*>
$$
$classes : Application \rightarrow Class*$ \\
$classes(<cs, is>) = cs, cs \in Class* \wedge is \in Instance* $ \\ \\
$class : Label \times Application \rightarrow Class  $ \\ 
$class(l, a) = c, c \in classes(a) \wedge name(c) = l, l \in Label, a \in Application* $ \\
%TODO: variable types

@@

$$Class = < label, Property*, Association* >$$
$name : Class \rightarrow Label$ \\
$name((lab, props, assocs)) = lab$ \\ \\
$properties : Class \rightarrow Property*$ \\
$properties((lab, props, assocs)) = props $\\ \\
$primitiveProperties : Class \rightarrow Property*$ \\
$primitiveProperties((lab, props, assocs)) = props', \forall p \in props' : cardinality(p) \leq 1 $\\ \\
$collectionProperties : Class \rightarrow Property*$ \\
$collectionProperties((lab, props, assocs)) = props', \forall p \in props' : cardinality(p) > 1 $\\ \\
$associations : Class \rightarrow Association*$ \\
$associations((lab, props, assocs)) = assocs $ \\ \\
$associated :  Class \times Application \rightarrow Boolean $
%TODO: variable types

\paragraph{Property} represents a feature of  a class which is can be represented as a primitive type. The property can be mandatory, can have a default value.
$$
Property = < label, AppType, DefaultValue, Cardinality, Mandatory >
$$
$name : Property \rightarrow Label$ \\
$name((lab, t, val, n)) = lab$ \\ \\
$type : Property \times AppType$ \\
$type((lab, t, val, n)) = t$ \\ \\
$cardinality : Property \rightarrow NzNat$ \\
$cardinality((lab, t, val, n)) = n$ \\ \\
$mandatory : Property \rightarrow Boolean $
$mandatory(<lab, t, val, n, m>) = m $ \\ \\
$owningClass : Property \times Application \rightarrow Class $ \\
$owningClass(p, a) = c, c \in classes(a) \wedge p \in properties(c) $
%TODO: variable types

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be a starting class of an association, referenced class is consider to be an ending class of an association. The cardinalities defines the multiplicity of the association.
$$
Association = (label, classRef, startCardinality, endCardinality)
$$
$name : Association \rightarrow Label$ \\
$name((lab, ref, n_1, n_2)) = lab$\\ \\
$reference : Association \rightarrow Label$ \\
$reference((lab, ref, n_1, n_2)) = ref$\\ \\
$startCardinality : Association \rightarrow NzNat$ \\
$startCardinality((lab, ref, n_1, n_2)) = n_1$\\ \\
$endCardinality : Association \rightarrow NzNat$ \\
$endCardinality((lab, ref, n_1, n_2)) = n_2$ \\ \\
$owningClass : Association \times Application \rightarrow Class $ \\
$owningClass(as, a) = c, c \in classes(a) \wedge as \in associations(c) $
%TODO: variable types

@@

\section{Database Model}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents instances. The database is a pair:
$$
Database = <table*, data*>
$$
$table : Label \times Database \rightarrow Table $ \\
$table(l, d) = t, t \in tables(d) \wedge name(t) = l$ \\

\subsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
$$
Table = (label, primaryKey, Column*, ForeignKey*)
$$
$name : Table \rightarrow Label $ \\
$name((lab, pk, cols, fks)) = lab$ \\ \\
$primaryKey : Table \rightarrow PrimaryKey $ \\
$primaryKey((lab, pk, cols, fks)) = pk$ \\ \\
$columns : Table \rightarrow Column* $ \\
$columns((lab, pk, cols, fks)) = cols$ \\ \\
$foreignKeys : Table \rightarrow ForeignKey* $ \\
$foreignKeys((lab, pk, cols, fks)) = fks$

\paragraph{Column} columns define possible data values and types which can be part of a table record.
$$
Column = (label, DbType, DefaultValue, Constraint*)
$$
$name : Column \rightarrow Label $ \\
$name((lab, t, val, cons)) = lab $ \\ \\
$type : Column \rightarrow DbType $ \\
$type((lab, t, val, cons)) = t $ \\ \\
$constraints : Column \rightarrow Constraint* $ \\
$constraints((lab, t, val, cons)) = cons $ \\ \\
$owningTable : Column \times Database \rightarrow Table $ \\
$constraints(c, d) = tab, tab \in tables(d) \wedge c \in columns(d) $

\paragraph{Foreign key}
$$
ForeignKey = (label, tableRef, Constraint*)
$$
$name : ForeignKey \rightarrow Label $ \\
$name((lab, tRef, cons)) = lab $ \\ \\
$reference : ForeignKey \rightarrow Label $ \\
$reference((lab, tRef, cons)) = tRef $ \\ \\
$reference : ForeignKey \rightarrow Constraint* $ \\
$reference((lab, tRef, cons)) = cons $ \\ \\
$owningTable : ForeignKey \times Database \rightarrow Table $ \\
$owningTable(fk, d) = tab, tab \in tables(d) \wedge fk \in foreignKeys(d) $

\paragraph{Primary key}
$$
PrimaryKey =  (label) 	
$$

@@

\subsection{Stored Data Model}

@@

$\rho_p(p, d) = \begin{cases}
  \inference{ d \vdash cardinality(p) = 1 \wedge ! mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d) 
 \\ \wedge \not instanciated(owningClass(p))
 }{ d \vdash columns(t) \cup column(name(p), \rho_t(type(p)), \varnothing )} \\ \\

  \inference{ d \vdash cardinality(p) = 1 \wedge mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d) 
 \\ \wedge ! instanciated(owningClass(p))
 }{ d \vdash columns(t) \cup column(name(p), \rho_t(type(p)), NOTNULL )} 
\\ \\

 \inference{d \vdash cardinality(p) > 1 \wedge ! mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}d \vdash  table(name(p), primaryKey("Id"), column("value", \rho_t(type(p))), \\ foreignKey(name(p), name(owningClass(p)), \varnothing) \end{gathered}}
\\ \\

 \inference{d \vdash cardinality(p) > 1 \wedge mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge ! instanciated(owningClass(p))}{\begin{gathered}d \vdash  table(name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), NOTNULL), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL) \end{gathered}}
\\
 \end{cases}$
\textbf{TODO: add value into instances in NOTNULL columns}

@@

$
\rho_a(a,d) = \begin{cases}
 \inference{d \vdash startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)}{d \vdash 
 \begin{gathered}
 foreignKeys(u) \\ \cup foreignKey(name(a), name(t),  \varnothing) 
 \end{gathered}
 }
  \\ \\
 \inference{d \vdash startCardinality(a) = 1 \\\\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)}{\begin{gathered} d \vdash 
foreignKeys(u) \\ \cup  foreignKey(name(a), name(t),  NOTNULL)
\end{gathered}
} \\ \\

  \inference{d \vdash  startCardinality > 1 \\ \wedge \rho(owningClass(p)) \in tables(d) \wedge \rho(reference(a)) \in tables(d)}{
  \begin{gathered} d \vdash 
 table(name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  \end{gathered}}  
 \end{cases}
$

@@

\begin{figure}
\begin{center}
	\includegraphics[scale=0.4]{./images/evolution}
	\caption{The evolution of data changes the system on all levels}
\end{center}
	\label{fig:evolution}
\end{figure}

@@

$
AddProperty(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in properties(p) : name(q) \neq name(p)}{
 properties(c) \cup p }
$
$$AddAssociation : Label \times Association \times Application \rightarrow Application $$
$
AddAssociation(l, s, a) = \inference{\forall f \in associations(a) : name(f) \neq name(s) \\ \wedge \exists c \in classes(a) : name(c) = l \\ \wedge \exists k \in classes(a) : name(k) = reference(s) }{associations(c) \cup s}
$

@@

$
\Phi_{addClass}(c, d) = \inference{d \vdash \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) 
}{
\begin{gathered}
d \vdash  tables(d) \cup \rho(c) \cup \rho(collectionProperties(c)) \\ \cup \rho(associations(c))
\end{gathered}
}
$
\paragraph{Add Property}
$
\Phi_{addProperty}(p, d) = 
\inference{d \vdash \exists t \in tables(d) : name(t) = name(owningClass(p))}{d \vdash \rho(p,d)} 
$
\paragraph{Add Association}
$
\Phi_{AddAssociation}(a, d) = \inference{d \vdash \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) }{d \vdash \rho(a,d)}
$

>>>>

$$
Application = <Class*>
$$
$classes : Application \rightarrow Class*$ \\
$classes(<cs, is >) = = cs, cs \in Class* \wedge is \in Instance* $ \\ \\
$class : Label \times Application \rightarrow Class  $ \\ 
$class(l, a) = c, c \in classes(a) \wedge name(c) = l, l \in Label, a \in Application* $ \\
%TODO: variable types

@@

$$Class = < label, Property*, Association* >$$
$name : Class \rightarrow Label$ \\
$name(< lab, props, assocs  >) = lab$ \\ \\
$properties : Class \rightarrow Property*$ \\
$properties(< lab, props, assocs  >) = props $\\ \\
$primitiveProperties : Class \rightarrow Property*$ \\
$primitiveProperties(< lab, props, assocs  >) = props', \forall p \in props' : cardinality(p) \leq 1 $\\ \\
$collectionProperties : Class \rightarrow Property*$ \\
$collectionProperties(< lab, props, assocs  >) = props', \forall p \in props' : cardinality(p) > 1 $\\ \\
$associations : Class \rightarrow Association*$ \\
$associations(< lab, props, assocs  >) = assocs $ \\ \\
$associated :  Class \times Application \rightarrow Boolean $
%TODO: variable types

\paragraph{Property} represents a feature of  a class which is can be represented as a primitive type. The property can be mandatory, can have a default value.
$$
Property = < label, AppType, DefaultValue, Cardinality, Mandatory >
$$
$name : Property \rightarrow Label$ \\
$name(< lab, t, val, n  >) = lab$ \\ \\
$type : Property \times AppType$ \\
$type(< lab, t, val, n  >) = t$ \\ \\
$cardinality : Property \rightarrow NzNat$ \\
$cardinality(< lab, t, val, n  >) = n$ \\ \\
$mandatory : Property \rightarrow Boolean $
$mandatory(<lab, t, val, n, m >) = = m $ \\ \\
$owningClass : Property \times Application \rightarrow Class $ \\
$owningClass(p, a) = c, c \in classes(a) \wedge p \in properties(c) $
%TODO: variable types

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be a starting class of an association, referenced class is consider to be an ending class of an association. The cardinalities defines the multiplicity of the association.
$$
Association = (label, classRef, startCardinality, endCardinality)
$$
$name : Association \rightarrow Label$ \\
$name(< lab, ref, n_1, n_2  >) = lab$\\ \\
$reference : Association \rightarrow Label$ \\
$reference(< lab, ref, n_1, n_2  >) = ref$\\ \\
$startCardinality : Association \rightarrow NzNat$ \\
$startCardinality(< lab, ref, n_1, n_2  >) = n_1$\\ \\
$endCardinality : Association \rightarrow NzNat$ \\
$endCardinality(< lab, ref, n_1, n_2  >) = n_2$ \\ \\
$owningClass : Association \times Application \rightarrow Class $ \\
$owningClass(as, a) = c, c \in classes(a) \wedge as \in associations(c) $
%TODO: variable types

@@

\section{Database Model}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents instances. The database is a pair:
$$
Database = <table*, row*>
$$
$table : Label \times Database \rightarrow Table $ \\
$table(l, d) = t, t \in tables(d) \wedge name(t) = l$ \\

\subsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
$$
Table = (label, primaryKey, Column*, ForeignKey*)
$$
$name : Table \rightarrow Label $ \\
$name(< lab, pk, cols, fks  >) = lab$ \\ \\
$primaryKey : Table \rightarrow PrimaryKey $ \\
$primaryKey(< lab, pk, cols, fks  >) = pk$ \\ \\
$columns : Table \rightarrow Column* $ \\
$columns(< lab, pk, cols, fks  >) = cols$ \\ \\
$foreignKeys : Table \rightarrow ForeignKey* $ \\
$foreignKeys(< lab, pk, cols, fks  >) = fks$

\paragraph{Column} columns define possible data values and types which can be part of a table record.
$$
Column = (label, DbType, DefaultValue, Constraint*)
$$
$name : Column \rightarrow Label $ \\
$name(< lab, t, val, cons  >) = lab $ \\ \\
$type : Column \rightarrow DbType $ \\
$type(< lab, t, val, cons  >) = t $ \\ \\
$constraints : Column \rightarrow Constraint* $ \\
$constraints(< lab, t, val, cons  >) = cons $ \\ \\
$owningTable : Column \times Database \rightarrow Table $ \\
$constraints(c, d) = tab, tab \in tables(d) \wedge c \in columns(d) $

\paragraph{Foreign key}
$$
ForeignKey = (label, tableRef, Constraint*)
$$
$name : ForeignKey \rightarrow Label $ \\
$name(< lab, tRef, cons  >) = lab $ \\ \\
$reference : ForeignKey \rightarrow Label $ \\
$reference(< lab, tRef, cons  >) = tRef $ \\ \\
$reference : ForeignKey \rightarrow Constraint* $ \\
$reference(< lab, tRef, cons  >) = cons $ \\ \\
$owningTable : ForeignKey \times Database \rightarrow Table $ \\
$owningTable(fk, d) = tab, tab \in tables(d) \wedge fk \in foreignKeys(d) $

\paragraph{Primary key}
$$
PrimaryKey =  < label > 	
$$

@@

\subsection{Model of Stored Data}
The representation of stored data in the relational database is based on pairs:
$$pair = < refColumn, value > | < refFK, value>$$

\textbf{TODO: primary key}
where the $refColumn$ reference a column or foreign key by its name and $value$ represents concrete value stored in the database. A row in a table is a set of pairs with reference to a concrete table:

$$row = < refTable, pair* >$$

\textbf{TODO: define consistency of rows and schema}

@@

$\rho_p(p, d) = \begin{cases}
  \inference{ d \vdash cardinality(p) = 1 \wedge ! mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d) 
 \\ \wedge \not instanciated(owningClass(p))
 }{\begin{gathered}
	 d \vdash columns(t) \cup column(name(p), \rho_t(type(p)), \varnothing )
\end{gathered}
 } \\ \\

  \inference{ d \vdash cardinality(p) = 1 \wedge mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}
 	d \vdash columns(t) \cup column(name(p), \rho_t(type(p)), NOTNULL ) \\

\end{gathered}
} 
\\ \\

 \inference{d \vdash cardinality(p) > 1 \wedge ! mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}d \vdash  table(name(p), primaryKey("Id"), column("value", \rho_t(type(p))), \\ foreignKey(name(p), name(owningClass(p)), \varnothing) \\

\end{gathered}}
\\ \\

 \inference{d \vdash cardinality(p) > 1 \wedge mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d) }{\begin{gathered}d \vdash  table(name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), NOTNULL), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL)
	 \end{gathered}}
 \end{cases}$

@@

$
\rho_a(a,d) = \begin{cases}
 \inference{d \vdash startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)}{d \vdash 
 \begin{gathered}
 foreignKeys(u) \\ \cup foreignKey(name(a), name(t),  \varnothing) 
 \end{gathered}
 }
  \\ \\
 \inference{d \vdash startCardinality(a) = 1 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)}{\begin{gathered} d \vdash 
foreignKeys(u) \\ \cup  foreignKey(name(a), name(t),  NOTNULL) \\
\end{gathered}
} \\ \\

  \inference{d \vdash  startCardinality > 1 \\ \wedge \rho(owningClass(p)) \in tables(d) \wedge \rho(reference(a)) \in tables(d)}{
  \begin{gathered} d \vdash 
 table(name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  \end{gathered}}  
 \end{cases}
$
\textbf{TODO: check + add cardinalities/images and split(?), data in NOTNULL}

@@

\begin{figure}
\begin{center}
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels}
\end{center}
	\label{fig:evolution}
\end{figure}

@@

$
AddProperty(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p }
$
$$AddAssociation : Label \times Association \times Application \rightarrow Application $$
$
AddAssociation(l, s, a) = \inference{\forall f \in associations(a) : name(f) \neq name(s) \\ \wedge \exists c \in classes(a) : name(c) = l \\ \wedge \exists k \in classes(a) : name(k) = reference(s) }{associations(c) \cup s}
$

@@

$
\Phi_{addClass}(c, d) = \inference{d \vdash \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) 
}{
\begin{gathered}
d \vdash  tables(d) \cup \rho(c) \cup \rho(collectionProperties(c)) \\ \cup \rho(associations(c))
\end{gathered}
}
$
\paragraph{Add Property}
$
\Phi_{addProperty}(p, d) = \begin{cases}
\inference{d \vdash \exists t \in tables(d) : name(t) = name(owningClass(p)) \wedge (! mandatory(p) \vee ! instantiated(owningClass(p))}{d \vdash \rho(p,d)} 
\\ \\ 
\inference{d \vdash \exists t \in tables(d) : name(t) = name(owningClass(p)) \wedge mandatory(p) \wedge instantiated(owningClass(p)) \wedge cardinality(p) \leq 1}{d \vdash \rho(p,d) \wedge \forall r \in data(d), reference(r) = t : pair(r) \cup pair(name(p), defaultValue(p))} 
\\ \\
\inference{d \vdash \exists t \in tables(d) : name(t) = name(owningClass(p)) \wedge mandatory(p) \wedge instantiated(owningClass(p)) \wedge cardinality(p) \leq 1}{\begin{gathered}
 \rho(p,d) \wedge \forall r \in row(owningClass(p)) : data(d)\\ \cup row(name(p), pair("value", defaultValue(p)), \\ pair(name(p), value(primaryKey(r))) 
\end{gathered}
}

\end{cases}
$

\paragraph{Add Association}
$
\Phi_{AddAssociation}(a, d) = \inference{d \vdash \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) }{d \vdash \rho(a,d)}
$
